---
title: 'Activity 03: Summary statistics, frequency distribution, and basic graphical
  representation'
author: "Javi Pascual"
date: "2026-06-01"
output: html_document
---

## Summary of Answers

### Most and least observed species
* **Most Observed Species:** Phlo (377 observations)
* **Least Observed Species:** Babu, Geat, Hety, Bosc, Drru,Apbo (1 observation each)

### Mean observations
         ASG     CVT ESG    OGF
Babu 0.00000000   0   0  0.04347826
Geat 0.00000000   0   0  0.04347826
Hety 0.06666667   0   0  0.00000000
Bosc 0.06666667   0   0  0.00000000
Drru 0.06666667   0   0  0.00000000
Apbo 0.00000000   0   0  0.04347826
Phlo 1.46666667   4   0 15.08695652

* **Habitat type with the highest species counts:** OGF (350 total observations)
* **Habitat types with no species observations:** None

################### INTERPRETATION #####################

  In the data from the article of Supsup et al. (2020) entitled "Variation in species richness, 
composition and herpetological community structure across a tropical habitat gradient of 
Palawan Island, Philippines", it can be seen that in terms of the total community observations 
across the dataset, Advanced Secondary Growth (ASG) had the highest amount with 58, which is then 
followed by Old Growth Forest (OGF, 48), Early Secondary Growth (ESG, 28), and Cultivation (CVT, 1),
which aligns with the aforementioned paper. However, the area with the most species counts is the OGF
with a significant different in amount of 350, followed by ASG (25), CVT (8), and ESG (0). It can then
be inferred that despite the amount of observations in ASG, it does not necessarily mean that it also hosts
the most species-rich area.

  The mean distribution provides insight into the dominant and specialized species in the different study sites.
The most observed species known aiss "Phlo" and the least observed species are Babu, Geat, Hety, Bosc, Drru, 
and Apbo. As for the Species Total Abundance (SPPTotal), it is seen in the scatterplot that it `most
apparent at lower elevations and would greatly decrease between 200 and 400 meters. In the histogram, 
it is heavily skewed with most of the obsercations found in the left part of the histogram. This can be
inferred to as the observed species to prefer the temperature or environment of that specific area as opposed
to the other areas.

################### END #####################

## Install and load packages rmarkdown and knitr ##
install.packages("rmarkdown")
install.packages("knitr")
install.packages("pander")
library(rmarkdown)
library(knitr)
library(pander)


```{r}

## Set working directory ##
setwd("C:/Users/Blight/Desktop/S.Y 25-26 Term 3/LBBBI27")
getwd()

```

```{r}

## Data cleaning ##
my_data <- read.csv("Data.csv")
remove_NA <- complete.cases(my_data)
head(my_data[remove_NA, ])
clean_data <- my_data[remove_NA, ]

```

```{r}

## Elevation ##
my_elevation <- clean_data$Elev

mean(my_elevation)

sd(my_elevation)

range(my_elevation)
```

```{r}

## Species data (G-AG) ##
my_speciesdata <- clean_data [ , 7:33]
print(my_speciesdata)
```

```{r}

## Typical loop ##
list_species <- list()

min_count <- Inf
max_count <- -1
min_species <- NULL
max_species <- NULL
current_name <- NULL
current_count <- -1
i <- 0

for (i in 1:27) {
  current_count <- sum(my_speciesdata[ , i])
  current_name <- names(my_speciesdata) [i]
  
  if (current_count < min_count) {
    min_count <- current_count
    min_species <- current_name
  } else if (current_count == min_count) {
    min_species <- c(min_species, current_name)
  }
  
  if (current_count > max_count) {
    max_count <- current_count
    max_species <- current_name
  } else if(current_count == max_count) {
    max_species <- c(max_species, current_name)
  }
  
  i <- i + 1
}

print("Least observations:")
print(min_species)
print(min_count)
# WILL PRINT
# [1] "Least observations:"
# [1] "Babu" "Geat" "Hety" "Bosc" "Drru" "Apbo"
# [1] 1
#

print("Most observations:")
print(max_species)
print(max_count)
# WILL PRINT
# [1] "Most observations:"
# [1] "Phlo"
# [1] 377



## Compact loop ##
species_counts <- sapply(my_speciesdata, sum)

min_spcounts <- min(species_counts)
max_spcounts <- max(species_counts)

min_spname <- names(species_counts[species_counts == min_spcounts])
max_spname <- names(species_counts[species_counts == max_spcounts])

print("Least observations:")
print(min_spname)
print(min_spcounts)
# WILL PRINT
# [1] "Least observations:"
# [1] "Babu" "Geat" "Hety" "Bosc" "Drru" "Apbo"
# [1] 1


print("Most observations:")
print(max_spname)
print(max_spcounts)
# WILL PRINT
# [1] "Most observations:"
# [1] "Phlo"
# [1] 377

```

```{r}

## Htypes ##
ASG_count <- 0
CVT_count <- 0
ESG_count <- 0
OGF_count <- 0
combine_htypes <- cbind(clean_data["Htypes"], clean_data["Babu"], clean_data["Geat"], clean_data["Hety"], clean_data["Bosc"], clean_data["Drru"], clean_data["Apbo"], clean_data["Phlo"])

mostsp_htmean <- tapply(combine_htypes$Phlo, combine_htypes$Htypes, mean)

split_ht <- split (combine_htypes, combine_htypes$Htypes)

ht_mean <- sapply(split_ht, function(x) {
    colMeans(x[ ,2:8])
})
  
ht_sum <- sapply(split_ht, function(x) {
    sum(x[,2:8])
})

print(ht_sum)
# WILL PRINT
# ASG CVT ESG OGF 
# 25   8   0 350 

```

```{r}
## Plots ##
plot_all <- clean_data[ , c("Htypes", "Elev", "SPPTotal")]

# Scatter plot #
plot(x = plot_all$Elev, y = plot_all$SPPTotal, xlab = "Elevation (meters)", ylab = "Species Total Count (SPPTotal)", main = "Species Total Count vs. Elevation", col = as.factor(plot_all$Htypes), pch = 19)

total_habitat <- tapply(plot_all$SPPTotal, plot_all$Htypes, sum)

# Bar plot #
barplot(total_habitat, xlab = "Habitat Type (Htypes)", ylab = "Total Species Observations", main = "Total Species Observed per Habitat Type", col = "orange")

# Histogram #
hist(plot_all$SPPTotal, xlab = "Species Total Counts (SPPTotal)", ylab = "Frequency (Number of Sites/Rows)", main = "Distribution of Total Species Counts", col = "orange", breaks = 12)

```

