---
title: 'Activity 4: Tidyverse'
author: "Javi Pascual"
date: "2026-06-01"
output: html_document
---

## Summary of answers
A.
5) IT2P7 and IT3P0 has the most count of small trees (both have 10)
Transect Htypes SmallTrees FloweringTrees
1    IT2P7    CVT         10              3
2    IT3P0    CVT         10              6
3    IT3P1    CVT          5              6
4    IT2P6    CVT          4              3
5    IT3P2    CVT          0              7
6    IT3P3    CVT          0              6
7    IT3P4    CVT          0             31

6) ASG and ESG do not differ significantly (5.26 and 5.38)
Htypes SL_mean
<chr>    <dbl>
1 ASG       5.26
2 CVT       8.86
3 ESG       5.38

7) Transects with no difference are: IT1P5, IT2P5, IT3P1, ST1P2, ST1P3. and more 
Transect Htypes SmallTrees FloweringTrees HerbGroundDiff
1     IT1P0    ESG         20             10            -10
2     IT1P1    ESG         15              3            -37
3     IT1P2    ESG         10              6            -55
4     IT1P3    ESG         15              5            -28
5     IT1P4    ESG         15              5            -10
6     IT1P5    ESG          4              4              0
7     IT1P6    ESG         16              4             40
8     IT2P0    ESG         13              4             50
9     IT2P1    ESG          3              3             35
10    IT2P2    ESG          4              3            -20
11    IT2P3    ESG          7              8             40
12    IT2P4    ESG         10              9             30
13    IT2P5    ESG         10              6              0
14    IT2P6    CVT          4              3             30
15    IT2P7    CVT         10              3             30
16    IT3P0    CVT         10              6            -15
17    IT3P1    CVT          5              6              0
18    IT3P2    CVT          0              7            -15
19    IT3P3    CVT          0              6            -58
20    IT3P4    CVT          0             31           -100
21    ST1P0    ASG         16              3            -20
22    ST1P1    ASG         10              2            -30
23    ST1P2    ASG          8              4              0
24    ST1P3    ASG         13              3              0
25    ST1P4    ASG         10              1            -30
26    ST1P5    ASG         11              3            -30
27    ST1P6    ASG         16              3            -10
28    ST1P7    ASG         10              1              0
29    ST1P8    ASG         13              1            -20
30    ST2P0    ASG          4              9             10
31    ST2P1    ASG          3              8             20
32    ST2P2    ASG          7              8             15
33    ST2P3    ASG          8              8             10
34    ST2P4    ASG          2              7            -10
35    ST2P5    ASG         10              6              0
36    ST2P6    ASG          0              6             -5
37    ST2P7    ASG          5              6            -15
38    ST2P8    ASG          7              9            -20
39    ST3P0    ASG          8              5             10
40    ST3P1    ASG          3              2              0
41    ST3P2    ASG          8              6              0
42    ST3P3    ASG          9              7            -35
43    ST3P4    ASG          9              6            -30
44    ST3P5    ASG         14              3            -30
45    ST3P6    ASG          6              9            -50
46    ST3P7    ASG          3             10            -70
47    ST3P8    ASG          3              6            -60


B. REFLECTION
How does using tidyverse compare to Activity 03, which primarily used base R? Which functions do you find more intuitive, and which ones are less so?
  Personally, using tidyverse is significantly easier than the previous use of base R. The functions are more intuitive and in tidyverse than base R. The overall usage
of tidyverse is much more straightforward and does not seem to be likely to be used with brute force coding due to its ease in usage.

## Setup tidyverse ##
install.packages("tidyverse")
library(tidyverse)


```{r chunk1, echo = TRUE}
## Set directory ##
setwd("C:/Users/Blight/Desktop/S.Y 25-26 Term 3/LBBBI27")
getwd()

```

```{r chunk2, echo = TRUE}
## For dataframe ##
library(tidyverse)
my_data <- read.csv("habitat_types.csv")

str(my_data)

my_data |> glimpse()

```

```{r chunk3, echo = TRUE}
## Subset, filter, and arrange data ##
subset1_mydata <- select(my_data, Transect, Htypes, SmallTrees, FloweringTrees)

filter_subset1 <- subset1_mydata[subset1_mydata$Htypes == "CVT",]

descen_subset1 <- arrange(filter_subset1, desc(SmallTrees))
print(descen_subset1)

```

```{r chunk4, echo = TRUE}
## Mean of flowering trees ##
mydata_formean <- group_by(my_data, Htypes)

mean_values <- summarise(mydata_formean, SL_mean = mean(FloweringTrees))
print(mean_values)

```

```{r chunk5, echo = TRUE}
## Adding the difference of herbscover and groundcover ##
self_df_mod <- subset1_mydata %>%
  mutate(HerbGroundDiff = my_data$HerbsCover - my_data$GroundCover)
print(self_df_mod)

```

```{r chunk6, echo = TRUE}
## Using pivot wider of self_df_mod ##
mod_wider <- self_df_mod |> pivot_wider(names_from = Htypes, values_from = c(SmallTrees, FloweringTrees, HerbGroundDiff))
print(mod_wider)

```














