---
title: "Pascual_linear_reg"
author: "Javi Pascual"
date: "2026-07-06"
output: html_document
---

```{r chunkA, echo=TRUE}
### Set working directory ###
setwd("C:/Users/Blight/Desktop/S.Y 25-26 Term 3/LBBBI27/Act 11_Linear_Regression")

### Read as dataframe ###
my_data <- read.csv("herpetozoa-033-095-s001.csv")

### Remove NAs ###
remove_NA <- complete.cases(my_data)
head(my_data[remove_NA, ])
clean_data <- my_data[remove_NA, ]

### Packages ##$
library(maps)
library(osmdata)
library(sf)
library(terra)
library(ggplot2)
library(dplyr)

```

```{r chunkB, echo=TRUE}
## Elevation  ##
ph_elev <- rast("ph_ strm_geo.tif")

## Subset longtitude and latitude ##
reduced_data <- clean_data[ , 2:3]

## Convert coordinates to spatial vector ##
spatvect_data <- vect(reduced_data)

## Extract elevation data ##
extracted_data <-  terra::extract(ph_elev, spatvect_data)

## Subset SPPTotal ##
subset_specrich <- clean_data[ , 6]

## Combine SPPTotal with extracted elevation ##
combined_data <-  cbind(subset_specrich, extracted_data)

## Rename ##
names(combined_data)[names(combined_data) == "Band_1"] <- "Elevation"

names(combined_data)[names(combined_data) == "subset_specrich"] <- "Species_Richness"

```

```{r chunkC, echo=TRUE}
## Simple linear regression plot  ##

# Independent variable #
independent_var <- combined_data$Elevation

# Dependent variable #
dependent_var <- combined_data$Species_Richness

## Mean ##
x_mean <- mean(independent_var)
y_mean <- mean(dependent_var)

## Slope ##

slope_reg <- sum((independent_var - x_mean) * (dependent_var - y_mean)) / sum((independent_var - x_mean)^2)

## Intercept ##
intercept_reg <- y_mean - slope_reg * x_mean

## Check v values ##
x_mean
y_mean
slope_reg
intercept_reg

## Equation ##
cat("Regression Equation:\n")
cat("y =", round(intercept_reg, 3), "+", round(slope_reg,3), "x\n")

## y hat ##
y_hat <- intercept_reg + slope_reg * independent_var

y_hat

data.frame(Elevation = independent_var, SpeciesRichness = dependent_var, Predicted = round(y_hat,2))

##  Plot with fitted line ##
plot(independent_var, dependent_var, pch = 19, col = "orange", xlab = "Elevation", ylab = "Species Richness", main = "Simple Linear Regression")

abline(intercept_reg, slope_reg, col = "aquamarine", lwd = 2)


## Plot with predicted values ##
plot(independent_var, dependent_var, pch = 19, col = "orange", xlab = "Elevation", ylab = "Species Richness", main = "Simple Linear Regression")

abline(intercept_reg, slope_reg, col = "aquamarine", lwd = 2)

## Predicted values ##
points(independent_var, y_hat, pch = 17,      # Triangle
       col = "pink", cex = 1.2)

## Residual lines ##
segments(x0 = independent_var, y0 = dependent_var, x1 = independent_var, y1 = y_hat, lty = 2, col = "lightgray")

## Legend ##
legend("topright", legend = c("Observed", "Predicted", "Regression Line"), pch = c(19, 17, NA), lty = c(NA, NA, 1), col = c("orange", "pink", "aquamarine"), lwd = c(NA, NA, 2), bty = "n")

## SSE, SST, and R squared ##
SSE <- sum((dependent_var - y_hat)^2)
SST <- sum((dependent_var - y_mean)^2)
R_squared <- 1 - (SSE / SST)

## Check values ##
SSE
SST
R_squared

```

What can you infer about the relationship between species and elevation?
A negative relationship can be observed between elevation and species richness. This is indicated by the regression line's orientation, which in this case is found to be proceeding downards as the graph moves to the right. Between the elevations of 600m to 1000m, the species richness is also found to significantly reduce, which can be inferred as to the possible limitation for the particular observed species.


```{r chunkD, echo=TRUE}
## Multiple linear regression ##

# Read as dataframe #
ph_precip <- rast("annual_precipitation.tif")
ph_temp <- rast("annual_temperature.tif")

## Extract data ##
extracted_precip <-  terra::extract(ph_precip, spatvect_data)
extracted_temp <-  terra::extract(ph_temp, spatvect_data)

## Check values ##
extracted_precip
extracted_temp

## Combine ##
combined_multi <-  combined_data |> mutate(Precipitation = extracted_precip [ ,2], Temperature = extracted_temp [,2])

## Fit regression model ##
reg_model <- lm(Species_Richness ~ Elevation + Precipitation + Temperature, data = combined_multi)

## Display model summary
summary(reg_model)

```

Which relationship is statistically significant?
Precipitation and temperature